---
title: "PROJECT"
author: "NITYA TANDON"
format: 
  pdf:
    warning: false
    message: false
    toc: false
    number-sections: false
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 2
editor: visual
---

```{r}
rm(list = ls())
options(scipen = 20)
```

```{r}
if (!require("jtools")) install.packages("jtools")
library(jtools)
if (!require("forcats")) install.packages("forcats")
library(forcats)
if (!require("tidyr")) install.packages("tidyr")
library(tidyr)
if (!require("dplyr")) install.packages("dplyr")
library(dplyr)
if (!require("expss")) install.packages("expss")
library(expss)
if (!require("stringr")) install.packages("stringr")
library(stringr)
if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)
if (!require("sandwich")) install.packages("sandwich")
library(sandwich)
if (!require("lmtest")) install.packages("lmtest")
library(lmtest)
if (!require("huxtable")) install.packages("huxtable")
library(huxtable)
if (!require("AER")) install.packages("AER")
library(AER)
if (!require("plm")) install.packages("plm")
library(plm)
if (!require("car")) install.packages("car")
library(car)
if (!require("stargazer")) install.packages("stargazer")
library(stargazer)
if (!require("clubSandwich")) install.packages("clubSandwich")
library(clubSandwich)
if (!require("broom")) install.packages("broom")
library(broom)

```

# LOADING AND CLEANING DATA:

```{r}
load("C:/Users/nitya/OneDrive/Desktop/PROJECT/nlsy97_long.Rda")
names(nlsy97_long)
```

## Making Birth year and finding out age

```{r}
nlsy97_long$birthyr <- as.numeric(nlsy97_long$birthyr)
nlsy97_long$year <- as.numeric(nlsy97_long$year)
nlsy97_long$age <- nlsy97_long$year - nlsy97_long$birthyr

summary(nlsy97_long$age)
```

## Creating the education variable and cleaning it and filled up

```{r}
nlsy97_long <- nlsy97_long %>% mutate(highested= case_when(
  year == 1997 ~ highestgrade,
  year >= 1998 & year <2010 ~ highested1,
  year >= 2010 ~ highested2))

summary(nlsy97_long$highested)

nlsy97_long <- nlsy97_long %>% mutate(highested= case_when(
  highested == 0 ~ NA,
  highested == 95 ~ NA,
  TRUE ~ highested))

summary(nlsy97_long$highested)

nlsy97_long <- nlsy97_long %>%
  arrange(personid, year) %>% 
  group_by(personid) %>%
  fill(highested, .direction = "down") %>% 
  ungroup()

summary(nlsy97_long$highested)
```

## Taking out missing varibles

```{r}
nlsy97_cleaned <- nlsy97_long %>%
  drop_na(highested) %>%
  filter(year != 2012 & year != 2014 & year != 2016 &
         year != 2018 & year != 2020 & year != 2022 & year != 2024)
```

## Table

```{r}
nlsy97_long %>% count(year)
nlsy97_cleaned %>% count(year)

aggregate(highested ~ age, data = nlsy97_cleaned, summary)

cor(nlsy97_cleaned$age, nlsy97_cleaned$highested)
```

## Box plot

```{r}
plot1 <- ggplot(nlsy97_cleaned, aes(as.factor(age), highested)) + 
  geom_boxplot(outliers = FALSE, fill = "white", colour = "#3366FF") +
  xlab("age") +
  ylab("Years of education obtained")

print(plot1)
```

# CREATING VARIABLES FOR WAGE AND INCOME AS WELL AS CONTROL VARIABLES

## Focusing on worker and hours worked, etc, with only looking at employeed workers with income and then creating a work income. Then calculating work income, wage and log wage

```{r}
nlsy97_cleaned <- nlsy97_cleaned %>% select(year, personid, workinc2, emphours, hadinc, workinc_est, everything()) 

nlsy97_cleaned <- subset(nlsy97_cleaned, hadinc==1 & !is.na(emphours) & emphours>0)

nlsy97_cleaned <- nlsy97_cleaned %>% mutate(workinc= case_when(
  workinc_est==1 ~ 2500,
  workinc_est==2 ~ 7500,
  workinc_est==3 ~ 17500,
  workinc_est==4 ~ 37500,
  workinc_est==5 ~ 75000,
  workinc_est==6 ~ 150000,
  workinc_est==7 ~ 300000,  
  TRUE ~ workinc2))

nlsy97_cleaned$wage <- nlsy97_cleaned$workinc/nlsy97_cleaned$emphours

nlsy97_cleaned <- nlsy97_cleaned %>% select(year, personid, wage, workinc, emphours, everything())

nlsy97_cleaned <- subset(nlsy97_cleaned, wage>0 & !is.na(wage)) 

nlsy97_cleaned$lnwage <- log(nlsy97_cleaned$wage)

nlsy97_cleaned <- nlsy97_cleaned %>% select(year, personid, lnwage, everything()) 

summary(nlsy97_cleaned$workinc)
summary(nlsy97_cleaned$wage)
summary(nlsy97_cleaned$lnwage)
```

## control variables = age square and age cube

```{r}
nlsy97_cleaned$agesq <- nlsy97_cleaned$age^2
nlsy97_cleaned$agecu <- nlsy97_cleaned$age^3

summary(nlsy97_cleaned$agesq)
summary(nlsy97_cleaned$agecu)
```

## Control variables = calculating tenure

```{r}
nlsy97_cleaned$tenure <- nlsy97_cleaned$tenure / 52
summary(nlsy97_cleaned$tenure)
```

## Control variable = Female variable

```{r}
nlsy97_cleaned$female <- as.numeric(nlsy97_cleaned$sex) - 1
summary(nlsy97_cleaned$female)
```

## Control ASVAB

```{r}
nlsy97_cleaned$asvab <- as.numeric(nlsy97_cleaned$asvab) / 1000
summary(nlsy97_cleaned$asvab)
```

## Non student age restriction

```{r}
 nlsy97_cleaned <- nlsy97_cleaned %>% mutate(student= case_when(
  year == 1997 ~ student2,
  year >= 1998 & year <2005 ~ student1,
  year >= 2005 ~ student2))

nlsy_agerestric <- subset(nlsy97_cleaned, student < 8 & age >= 30)

nlsy_agerestric$experience_l <- nlsy_agerestric$age - nlsy_agerestric$highested - 6

summary(nlsy_agerestric$experience_l)
```

# REGRESSION 1: the baseline regression

```{r}
reg1 <- lm(
  lnwage ~ highested + age + agesq + agecu + union + tenure + female + as.factor(year),
  data = nlsy_agerestric
)

summary(reg1)

reg1a <- coeftest(reg1, vcov = vcovHC, type = "HC0")
reg1b <- coeftest(reg1, vcov = vcovCL, type = "HC0", cluster = ~personid)

export_summs(
  reg1, reg1a, reg1b,
  model.names = c("Basic", "Robust SE", "Clustered SE"),
  digits = 4,
  coefs = c("highested", "age", "agesq", "agecu", "union", "tenure", "female")
)
```

# REGRESSION 2: the baseline regression with asvab as a control

```{r}
reg2 <- lm(
  lnwage ~ highested + asvab + age + agesq + agecu + union + tenure + female + as.factor(year),
  data = nlsy_agerestric
)

summary(reg2)

reg2a <- coeftest(reg2, vcov = vcovCL, type = "HC0", cluster = ~personid)

export_summs(
  reg1b, reg2a,
  model.names = c("Education", "+ASVAB"),
  digits = 4,
  coefs = c("highested", "asvab", "age", "agesq", "agecu", "union", "tenure", "female")
)
```

# LOADING AND CLEANING FIELD OF STUDY

## Adding the college majors

```{r}
majors <- read.table(
  "C:/Users/nitya/OneDrive/Desktop/PROJECT/projectmajorssmaller.dat",
  header = FALSE, sep = " "
)
names(majors)
```

## Loading names of the data in the NLSY97 as is to identify

```{r}
colnames(majors) <- c(
  "personid",
  "major2013_01", "major2013_02", "major2013_03", "major2013_04",
  "major2015_01", "major2015_02", "major2015_03", "major2015_04",
  "major2017_01", "major2017_02", "major2017_03", "major2017_04", "major2017_05",
  "major2019_01", "major2019_02", "major2019_03", "major2019_04", "major2019_05",
  "major2021_01",
  "major2023_01", "major2023_02", "major2023_03"
)
```

## Extracting data

```{r}
majors <- majors %>% mutate(across(everything(), as.numeric))

majors_clean <- majors %>%
  mutate(across(-personid, ~ case_when(
    .x < 0  ~ NA_real_,
    .x == 999 ~ NA_real_,
    TRUE      ~ .x
  )))

majors_clean <- majors_clean %>%
  mutate(
    major_raw = coalesce(
      major2013_01, major2013_02, major2013_03, major2013_04,
      major2015_01, major2015_02, major2015_03, major2015_04,
      major2017_01, major2017_02, major2017_03, major2017_04, major2017_05,
      major2019_01, major2019_02, major2019_03, major2019_04, major2019_05,
      major2021_01,
      major2023_01, major2023_02, major2023_03
    )
  )

majors_final <- majors_clean %>% select(personid, major_raw)

sum(!is.na(majors_final$major_raw))
table(majors_final$major_raw, useNA = "ifany")
```

## Mapping the numbers into field of study groups

```{r}
majors_final <- majors_final %>%
  mutate(
    field_group = case_when(
      major_raw %in% c(11, 14, 15, 26, 27, 40, 41) ~ "STEM",
      major_raw %in% c(52) ~ "Business",
      major_raw %in% c(42, 43, 44, 45) ~ "Social Science/Public Service",
      major_raw %in% c(16, 23, 38, 39, 50, 54, 5, 9) ~ "Humanities/Arts",
      major_raw %in% c(13, 32, 33) ~ "Education",
      major_raw %in% c(34, 51, 60) ~ "Health",
      major_raw %in% c(10, 12, 46, 47, 48, 49) ~ "Trades/Applied",
      major_raw %in% c(1, 3, 4, 19, 24, 25, 28, 29, 30, 31, 35, 36, 37, 53, 96) ~ "Other",
      TRUE ~ NA_character_
    )
  )

table(majors_final$field_group, useNA = "ifany")
```

## Merging the majors and field of study part into the nlsy agerestrict

```{r}
nlsy_agerestric   <- nlsy_agerestric   %>% mutate(personid = as.numeric(personid))
majors_final <- majors_final %>% mutate(personid = as.numeric(personid))

nlsy_final <- nlsy_agerestric %>% left_join(majors_final, by = "personid")

nrow(nlsy_final)
sum(!is.na(nlsy_final$field_group))
```

# CREATING A DEGREE VARIABLE

```{r}
degree_max <- nlsy_final %>%
  mutate(highested3 = as.numeric(highested3)) %>%
  filter(highested3 >= 0) %>%
  group_by(personid) %>%
  summarise(degree_raw = max(highested3, na.rm = TRUE)) %>%
  ungroup()

nlsy_final <- nlsy_final %>%
  select(-any_of("degree_raw")) %>%
  left_join(degree_max, by = "personid")

nlsy_final <- nlsy_final %>%
  mutate(degree_type = case_when(
    degree_raw %in% c(0, 1) ~ "Less than HS / GED",
    degree_raw == 2         ~ "High School Diploma",
    degree_raw == 3         ~ "Associate",
    degree_raw == 4         ~ "Bachelor's",
    degree_raw == 5         ~ "Master's",
    degree_raw == 6         ~ "PhD",
    degree_raw == 7         ~ "Professional",
    TRUE                    ~ NA_character_
  ))

table(nlsy_final$degree_type, useNA = "ifany")
```

# CREATING A POST SECONDARY EDUCATION DISTINCTION VARIABLE

```{r}
nlsy_dergee <- nlsy_final %>%
  filter(
    !is.na(field_group),
    degree_type %in% c("Associate", "Bachelor's", "Master's", "PhD", "Professional")
  )

nrow(nlsy_dergee)
length(unique(nlsy_dergee$personid))
table(nlsy_dergee$field_group)
table(nlsy_dergee$degree_type)
```

# FIG 1: MEAN LOG WAGE BY LEVEL OF DEGREE ACHIEVED

```{r}
deg_means <- nlsy_final %>%
  filter(!is.na(degree_type)) %>%
  group_by(degree_type) %>%
  summarise(mean_lnwage = mean(lnwage, na.rm = TRUE), n = n()) %>%
  mutate(degree_type = factor(degree_type,
    levels = c("Less than HS / GED", "High School Diploma", "Associate",
               "Bachelor's", "Master's", "PhD", "Professional")))

fig1 <- ggplot(deg_means, aes(x = degree_type, y = mean_lnwage, fill = degree_type)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  scale_fill_brewer(palette = "Blues") +
  xlab("Highest degree earned") +
  ylab("Mean log hourly wage") +
  ggtitle("Figure 1: Mean Log Wage by Degree Level (base)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

print(fig1)
```

# FIG 2: MEAN LOG WAGE BY TYPE OF STUDY

```{r}
field_means <- nlsy_dergee %>%
  group_by(field_group) %>%
  summarise(mean_lnwage = mean(lnwage, na.rm = TRUE), n = n()) %>%
  arrange(mean_lnwage)

fig2 <- ggplot(field_means,
               aes(x = mean_lnwage,
                   y = fct_reorder(field_group, mean_lnwage))) +
  geom_point(color = "#3366FF", size = 3) +
  geom_segment(aes(xend = min(mean_lnwage) - 0.02, yend = field_group),
               color = "#3366FF", linewidth = 0.4) +
  xlab("Mean log hourly wage") +
  ylab("") +
  ggtitle("Figure 2: Mean Log Wage by Field of Study (base)") +
  theme_bw() +
  theme(panel.grid.major.y = element_blank())

print(fig2)
```

# REGRESSION 3: WAGE BY LEVEL OF DEGREE ACHIEVED

```{r}
nlsy_final$degree_type <- relevel(factor(nlsy_final$degree_type), ref = "High School Diploma")

reg3 <- lm(
  lnwage ~ degree_type + age + agesq + agecu + union + tenure + female + as.factor(year),
  data = nlsy_final
)

summary(reg3)
reg3a <- coeftest(reg3, vcov = vcovCL, type = "HC0", cluster = ~personid)

```

# REGRESSION 4: WAGE BY LEVEL OF DEGREE ACHIEVED WITH ASVAB AS AN ABILITY CONTROL

```{r}
reg4 <- lm(
  lnwage ~ degree_type + asvab + age + agesq + agecu + union + tenure + female + as.factor(year),
  data = nlsy_final
)

reg4a <- coeftest(reg4, vcov = vcovCL, type = "HC0", cluster = ~personid)

summary(reg4)
```

## Table: Degrees with and without ASVAB

```{r}
export_summs(
  reg3a, reg4a,
  model.names = c("Degree (no ASVAB)", "Degree + ASVAB"),
  digits = 4,
  coefs = c("degree_typeAssociate", "degree_typeBachelor's",
            "degree_typeMaster's", "degree_typePhD",
            "degree_typeProfessional", "degree_typeLess than HS / GED",
            "asvab", "age", "agesq", "agecu", "union", "tenure", "female")
)
```

# REGRESSION 5: WAGES BY FIELD OF STUDY

```{r}
nlsy_dergee$field_group <- relevel(factor(nlsy_dergee$field_group),
                                 ref = "Social Science/Public Service")

reg5 <- lm(
  lnwage ~ field_group + age + agesq + agecu + union + tenure + female + as.factor(year),
  data = nlsy_dergee
)

summary(reg5)

reg5a <- coeftest(reg5, vcov = vcovCL, type = "HC0", cluster = ~personid)
```

# REGRESSION 6: WAGES BY FIELD OF STUDY + DEGREE ACHIEVED

```{r}
nlsy_dergee$degree_type <- relevel(factor(nlsy_dergee$degree_type), ref = "Bachelor's")

reg6 <- lm(
  lnwage ~ field_group + degree_type + age + agesq + agecu + union + tenure + female + as.factor(year),
  data = nlsy_dergee
)

summary(reg6)

reg6a <- coeftest(reg6, vcov = vcovCL, type = "HC0", cluster = ~personid)
```

## Table: FIELD OF STUDY WITH AND WITHOUT DEGREE AS A CONTROL

```{r}
export_summs(
  reg5a, reg6a,
  model.names = c("Field only", "Field + Degree"),
  digits = 4,
  coefs = c("field_groupSTEM", "field_groupBusiness",
            "field_groupHealth", "field_groupHumanities/Arts",
            "field_groupEducation", "field_groupTrades/Applied", "field_groupOther",
            "degree_typeAssociate", "degree_typeMaster's",
            "degree_typePhD", "degree_typeProfessional",
            "union", "tenure", "female")
)
```

# REGRESSION 7: WITH ASVAB AS AN ABILITY CONTROL

```{r}
reg7 <- lm(
  lnwage ~ field_group + degree_type + asvab + age + agesq + agecu + union + tenure + female + as.factor(year),
  data = nlsy_dergee
)

summary(reg7)

reg7a <- coeftest(reg7, vcov = vcovCL, type = "HC0", cluster = ~personid)
```

## TABLE: FIELD AND DEGREES BEOFRE AND AFTER ADDING ASVAB AS A CONTROL

```{r}
export_summs(
  reg6a, reg7a,
  model.names = c("Field + Degree (no ASVAB)", "Field + Degree + ASVAB"),
  digits = 4,
  coefs = c("field_groupSTEM", "field_groupBusiness",
            "field_groupHealth", "field_groupHumanities/Arts",
            "field_groupEducation", "field_groupTrades/Applied", "field_groupOther",
            "asvab", "union", "tenure", "female")
)
```

# Figure 4: Coefficient plot — field premiums with 95% CI

```{r}
field_coefs <- tidy(reg6a) %>%
  filter(grepl("field_group", term)) %>%
  mutate(
    field = gsub("field_group", "", term),
    lower = estimate - 1.96 * std.error,
    upper = estimate + 1.96 * std.error
  )

fig4 <- ggplot(field_coefs,
               aes(x = estimate,
                   y = fct_reorder(field, estimate),
                   color = estimate > 0)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("TRUE" = "#3366FF", "FALSE" = "#CC3300"),
                     guide = "none") +
  xlab("Coefficient (ref: Social Science/Public Service)") +
  ylab("") +
  ggtitle("Figure 4: Field of Study Wage Premiums\n(95% CI, conditional on degree, age, controls)") +
  theme_bw()

print(fig4)
```

# Figure 5: STEM and Health premium — with vs. without ASVAB

```{r}
stems <- data.frame(
  field = rep(c("STEM", "Health"), 2),
  model = c("No ASVAB", "No ASVAB", "With ASVAB", "With ASVAB"),
  est   = c(coef(reg6)["field_groupSTEM"],
            coef(reg6)["field_groupHealth"],
            coef(reg7)["field_groupSTEM"],
            coef(reg7)["field_groupHealth"]),
  se    = c(sqrt(vcov(reg6)["field_groupSTEM",  "field_groupSTEM"]),
            sqrt(vcov(reg6)["field_groupHealth", "field_groupHealth"]),
            sqrt(vcov(reg7)["field_groupSTEM",  "field_groupSTEM"]),
            sqrt(vcov(reg7)["field_groupHealth", "field_groupHealth"]))
)

fig5 <- ggplot(stems, aes(x = model, y = est, fill = model)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_errorbar(aes(ymin = est - 1.96 * se, ymax = est + 1.96 * se), width = 0.15) +
  facet_wrap(~field) +
  scale_fill_manual(values = c("No ASVAB" = "#CC3300", "With ASVAB" = "#3366FF")) +
  xlab("") +
  ylab("Field premium (log points)") +
  ggtitle("Figure 5: STEM and Health Premiums Before and After Controlling for ASVAB") +
  theme_bw()

print(fig5)

```
